\section{Measurements}

I evaluated the runtime overhead of my approach using the SPEC
CPU2006~\cite{spec-cpu2006} suite of C/\CPP and Fortran benchmarks. SPEC's
target config files~\cite{spec-cpu2006-config}, along with a few modifications
and additions in their global \texttt{Makefile.defaults} file, allow the
benchmarks to be conveniently built without any further changes both by Clang,
with or without CFCSS instrumentation, and by GCC for comparison, passing the
desired configuration to \texttt{runspec} via the \texttt{--config=...} option.

CPU2006 provides the \texttt{OBJOPT} variable and its language-specific
variants \texttt{COBJOPT}, \texttt{CXXOBJOPT} and \texttt{FOBJOPT} \cite[2.
Items used in Makefile.defaults]{spec-cpu2006-makevars} for the user to specify
in a target config file how her compiler can be instructed to give an object
file an arbitrary name, e.g. \texttt{-o \textit{filename}} in the case of both
GCC and Clang. \todo{long \& complicated sentence}Although it might slightly
obscure intent, I use \texttt{OBJOPT} to also pass \texttt{-emit-llvm} to
\texttt{clang}/\texttt{clang++} in my configurations for Clang with and without
CFCSS instrumentation—the latter to ensure that compilation results end up
roughly the same in both cases—in order for the object files to contain LLVM
bitcode after this step.

I further introduce a \texttt{CFCSS\_INSTRUMENTATION} boolean. In the case of
it being defined, \texttt{Makefile.defaults} will first run
\todo{citation}\texttt{llvm-link} to produce a single bitcode module of all
object files handed to it. It will then run the \texttt{opt} utility with the
following set of passes\footnote{Documentation for \texttt{simplifycfg} and
\texttt{mem2reg} can be found in \cite{llvm-passes}.}:

\begin{description}
  \item[simplifycfg] \cite[simplifycfg]{llvm-passes} to preprocess the module,
    since dangling basic blocks without predecessors (encountered once during
    trials of compiling \texttt{libgit2}, for reasons unknown) would throw of
    instrumentation, see \todo{link}BUG
  \item[instrument-blocks] if \texttt{CFCSS\_INSTRUMENTATION} is true, to
    perform CFCSS instrumentation
  \item[mem2reg] to promote local variable references introduced in the
    previous step to virtual registers \cite[mem2reg]{llvm-passes}; this
    \todo{check \& update}should be a no-op if \texttt{CFCSS\_INSTRUMENTATION}
    was set to \texttt{false}.
  \item[simplifycfg] \todo{reference}to cancel the effects of previously
    running SplitAfterCall as part of InstrumentBlocks; this again should be
    a no-op if CFCSS instrumentation was not requested
\end{description}

\texttt{gfortran} is still used in the Clang case to \todo{citation as to why
LLVM has no own front end for this}parse Fortran code, emitting LLVM
bitcode\footnote{Due to limitations in using GCC with DragonEgg, the
\texttt{-S} option has to be used with GCC in this case and LLVM bitcode is
emitted in human-readable form~\cite[DragonEgg in action]{dragonegg}. Although
this might slow down compilation a little, the overhead is assumed to be
negligible. Plus if you are using Fortran you probably do not want CFCSS
anyway.} via the DragonEgg GCC plugin \cite{dragonegg} which is then
instrumented and turned into a binary like the C/\CPP benchmarks.

\subsection{Performance}

\begin{itemize}
  \item instructions added / code size overhead
  \item common case IPC performance on Fiasco (Ping-Pong)
  \item SPEC on Linux
\end{itemize}

\subsection{Fault Detection}

To verify the expected error detecting capabilities of my CFCSS implementation,
I developed a tool for the injection of control flow errors, using Intel®
Pin~\cite{intel-pin}. The classes of faults that can be injected are the ones
mentioned in \cite[IV. Simulation Results]{oh-2002-control}.

Branch instructions at a given address can be deleted, equivalent to being
replaced with a NOP instruction. Unconditional branch instructions can be
inserted at a given address, provided it did refer to a valid instruction of
the original program before. Branch operand changes are implemented by first
verifying that the instruction at a given address is indeed a direct branch or
call, taking the target address, picking a random bit within it and flipping
it, and lastly inserting a new, direct branch to the resulting address in place
of the original instruction.

The tool allows limiting the target of any newly inserted branches to
executable sections of the program, though not necessarily addresses which did
previously mark the beginning of an instruction. Since branches to other memory
regions would immediately result in a segmentation fault anyway, they are not
that relevant for the testing of CFCSS functionality. Performing a new bit flip
on the original address instead allows the tool to exercise more interesting
cases without delay, although one has to be aware that the relative frequencies
of certain error classes will be skewed.

\begin{itemize}
  \item in relation to measurements from original paper
\end{itemize}

\tobewritten[inline]{fill}
